A Large and Variable Leading Tail of Helium in a Sub-Saturn Undergoing Runaway Inflation

### 2023-06-26 paper/paper1/ms.tex
Up to page 6 now

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2000ApJ...537..916S,2003Natur.422..143V,2012ApJ...751...86J,2018NatAs...2..714Y,2018ApJ...855L..11O,2018Natur.557...68S, 2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ 3470 b} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  Its $\sim$0.05 g~cm$^{-3}$ density places it below the 0.1 g~cm$^{-3}$ threshold predicted to exhibit runaway inflation by \citet{2023ApJ...945L..36T}.  Figures \ref{fig:instellation} and \ref{fig:ThornFortUpdated} show how much of an outlier \hatpb~ is: large, low mass, and heavily irradiated.

The evolutionary state of HAT-P-67~b offers even more intriguing possibilities.  The host star may be an evolved subgiant, opening the prospect for planetary re-inflation \citep{2022AJ....163...53S, 2022AJ....163..120G, 2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase in this part of the short-lived HR diagram.  The tidal gravity of the nearby massive host star may amplify mass loss rates even further \citep{2007A&A...472..329E, 2023ApJ...945L..36T}. Its status as a rare inflated Hot Saturn makes HAT-P-67~b a promising laboratory, uniquely suited for testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis}, we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in Section \ref{secHeAnalysis}, with deep dives on the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with 1D Parker wind models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

@@

We can constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i}$, and the observation of $i\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the sub-giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

We conclude that reinflation likely does not have a significant effect on the evolution of HAT-P-67~b: inflation timescales remain relatively unchanged over the planet lifetime, despite a 50\% spike in incident radiation in the sub-giant scenario.  The reason is subtle, as we describe next.

>>>>

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2000ApJ...537..916S,2003Natur.422..143V,2012ApJ...751...86J,2018NatAs...2..714Y,2018ApJ...855L..11O,2018Natur.557...68S, 2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ 3470 b} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  The sample of detections includes both hot Jupiters and lower mass planets, but lacks inflated hot Saturns owing to their intrinsic rarity below the 0.1 g~cm$^{-3}$ threshold. An observational picture of mass loss in inflated hot Saturns appears to be lacking for this reason.

@@

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, a transiting inflated hot Saturn orbiting an F5 sub giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity ($\log{g_p}<2.3$, cgs) makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  Its $\sim$0.05 g~cm$^{-3}$ density places it below the 0.1 g~cm$^{-3}$ threshold predicted to exhibit runaway inflation by \citet{2023ApJ...945L..36T}.  Figures \ref{fig:instellation} and \ref{fig:ThornFortUpdated} show how much of an outlier \hatpb~ is: large, low mass, and heavily irradiated.

The evolutionary state of HAT-P-67~b offers even more intriguing possibilities.  The planet may have undergone re-inflation \citep{2022AJ....163...53S, 2022AJ....163..120G, 2023arXiv230306728G}, as the host star evolved through the subgiant phase-- with the planet's insolation increasing with the star's rapid luminosity increase in this part of the short-lived HR diagram.  The tidal gravity of the nearby massive ($M_\star \sim1.6\;M_\odot$) host star may amplify mass loss rates even further \citep{2007A&A...472..329E, 2023ApJ...945L..36T}. Its status as a rare inflated hot Saturn makes HAT-P-67~b a promising laboratory, uniquely suited for testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis}, we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in Section \ref{secHeAnalysis}, with analysis of the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with 1D Parker wind models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

\subsection{Gaia DR3}\label{gaiadr3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc.  Minor corrections to the parallax uncertainty \citep{2021MNRAS.506.2269E} and bias \citep{2021A&A...649A...4L} appear negligible for the $G=9.98$ mag source.  The DR3 parallax places the system about 8.7\% farther than previously estimated by \citet{2017AJ....153..211Z}, which adopted a \emph{Gaia} DR1-informed parallax of $2.92\pm0.23$ mas, including a systematic $-0.325$ mas bias term \citep{2016ApJ...831L...6S}.  The wide companion HAT-P-67B has nearly identical parallax ($2.58\pm0.05$ mas) and proper motions, confirming its interpretation as co-moving, with a projected separation of about 3400 AU.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet, which orbits the primary.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.

@@

We can constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i_\star}$, and the observation of the planet's orbital inclination $i_p\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We assume spin-orbit alignment, $i_\star \sim i_p$.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the sub giant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

We conclude that reinflation likely does not have a significant effect on the evolution of HAT-P-67~b: inflation timescales remain relatively unchanged over the planet lifetime, despite a 50\% spike in incident radiation in the sub giant scenario.  The reason is subtle, as we describe next.
